Fast and Flexible Binary Instrumentation for the Kernel

### 2013-12-10 VEE-14/paper.tex
Minor edits to sections 1 .. 3. Fixed typos, grammar, and flow issues.

Instrumenting all code (e.g., the whole kernel) imposes unnecessary performance overheads when only a subset of the code is being analyzed (e.g., modules, functions). In practice, we would like to target instrumentation at specific code for efficiency, while also allowing rich and detailed instrumentation of the targeted code. Limiting the scope of what is instrumented is  important because the overhead introduced by using runtime code instrumentation is usually dominated by the instrumentation itself and not the baseline overhead of the runtime system supporting that instrumentation. Also, specializing based on what is instrumented allows instrumentation tools focus instrumentation on the code of interest. For example, if one wants to instrument RCU read-side and write-side critical sections, as is the case with Granary's \toolname{RCUdbg} tool, then instrumenting all code just to filter out / find those sections is wasteful.

@@

Another limitation of existing systems is that they are unable to perform efficient runtime code specialization. That is, they are unable to change how code is instrumented based on the context in which that code will run. This is unfortunate because it means that tools wanting to instrument code differently must create their own solution that works around this deficiency. For example, one way that existing tools can specialize code is to generate as many versions of the code as there are potential specializations. When instrumented code is executing, runtime checks continually check and jump to the correct version of the code for the current execution context. However, runtime checks are inefficient, and generating many versions of the code introduces bloat and pollutes the CPU's instruction cache. Also, the runtime checks needed to multiplex between different versions of instrumented code is orthogonal to the true goals of the instrumentation tool, thus distracting tool-developers from their primary objectives.

@@

One example of a tool that uses runtime code specialization is Granary's Read-Copy-Update (RCU) debugging tool, \toolname{RCUdbg}. \toolname{RCUdbg} catches bugs related to invalid uses of RCU-protected pointers (\Cref{sec:rcudbg}). \toolname{RCUdbg} distinguishes between two kinds of RCU-protected pointers (reader-side and writer-side) and plain old pointers, and three execution contexts (inside of a read-side critical section, inside of a write-side critical section, and in non-RCU-related code). The actions taken when a pointer is de-referenced varies depending on the kind of pointer and the execution context. Code executing in read-side critical sections is specialized to enforce that only plain-old and read-side pointers are accessed, and that all accesses through read-side pointers are memory reads. Write-side critical sections are specialized in a similar way.

@@

Granary translates and instruments kernel binaries one basic block at a time. In Granary, a basic block is a sequence of x86-64 instructions ending in a conditional branch, \texttt{ret}, or a \texttt{jmp} instruction. Unlike DRK and DynamoRIO, Granary does not terminate basic blocks at \texttt{call} instructions. This enables longer basic blocks and native-like proceccor return address prediction, but makes Granary non-transparent (\Cref{sec:transparency}). Entrypoint basic blocks (e.g. interrupt vectors, system call entrypoints, module initializers) are translated ahead-of-time, and all other basic blocks are translated just-in-time (JIT) as execution ``discovers'' them.

@@

Granary's just-in-time translation approach means that code executing from the code cache may yield control to Granary to request the address of the next basic block to execute. When instrumented code yields to Granary, a ``context switch'' occurs that transfers execution to a CPU-private stack where Granary operates. Granary operates on a CPU-private stack for two reasons. First, if Granary did not switch stacks, then context switches into Granary risk overflowing the runtime call stack. This can happen because kernel stacks are typically small (4K to 8K in size), and because a context switch could happen when the native stack is nearly exhausted. Second, on an SMP, multiple processors can simultaneously execute Granary code (e.g. to translate code that will later execute). In such cases, Granary depends on CPU-private data structures to avoid contention, and the stack is one such data structure. A context-switch back to the code cache occurs when the next basic block has been found or translated. At this point, instrumented execution continues. Similar to other DBT systems, Granary uses caching and hot code patching to reduce the number of context switches. In our experience, kernel code stabilizes very quickly: context switches stop happening after the first few seconds of translation.

@@

Associated with each basic block is meta-data that describes the basic block. Basic block meta-data includes: \begin{itemize}
	\item The length in bytes of the original and translated basic blocks, and the address of the first instruction in the original basic block and the instrumented basic block. This information helps provide a rich debugging interface for instrumentation tool authors. Granary is unique among DBT tools in that the debugging experience is actually pleasant. For example, Granary has a built-in debugging mechanism for stepping back in time through previously executed basic blocks, as well as querying past register state.
	\item Linux kernel-specific exception table meta-data about instructions contained within the current basic block. This kernel-specific meta-data is used by Granary to implement transparent page fault exception handling (\Cref{sec:interrupt_return_address_transparency}).
	\item Ranges of instructions that must be executed atomically. Granary uses this information to implement selective interrupt delaying (\Cref{sec:interrupt_delay}).
	\item The allocator used to allocate the instructions of the basic block. Granary uses this information to implement its trace-allocation optimization (\Cref{sec:trace_alloc}).
	\item The instrumentation policy used to instrument the basic block. Granary uses this information to allow instrumentation tools to define policy-specific interrupt handling routines. For example, \toolname{RCUdbg} implements a read-side policy-specific interrupt handling routine that reports an error if a code executing within a read-side critical section is interrupted (\Cref{sec:rcudbg}).
\end{itemize} %
%Granary uses basic block meta-data for trace allocation (\Cref{sec:trace_alloc}), interrupt delaying (\Cref{sec:interrupt_delay}), and policy-specific interrupt handling.

Basic block meta-data can be also be extended by instrumentation tools. This makes developing instrumentation tools that record per-basic block statistics very easy to make. For example, Granary's control-flow graph tool, \toolname{cfg}, extends the basic block meta-data to track per-block and per-edge execution counts (\Cref{sec:cfg}). This data is useful because it provides precise code coverage information (block and branch coverage), as well as useful data for profile-guided optimization. This meta-data is specified by \toolname{cfg} and automatically allocated for \toolname{cfg} by Granary.

@@

\subsection{Transparency}\label{sec:transparency}
An instrumentation system is transparent if, given the same inputs, the instrumented and uninstrumented versions of a program behave in the same way \cite{Transparency}. Granary includes configurable levels of transparency to address the trade-off between transparency and overhead. By default, Granary uses a relaxed transparency model for efficiency, flexibility and increased visibility, but if instrumentation for some module requires transparency then it can be enabled at the cost of increased overheads and decreased visibility. By default, Granary exposes the following two artifacts: code cache addresses as function and interrupt return addresses.

@@

\subsubsection{Granary}
For efficiency, most kernel-space instrumentation systems (Granary included) operate on CPU-private state (e.g. Granary's CPU-private stacks). However, accessing CPU-private without first disabling interrupts results in undefined behavior. Therefore, context-switches into Granary first disable interrupts and then switch execution onto one of Granary's CPU-private stacks. Context switches out of Granary and into native code (\Cref{sec:detach}) or the code cache returns execution to the native stack and re-enable interrupts. Execution behavior is preserved because Granary saves the machine state (i.e. registers, flags) in a reentrant way. All other parts of Granary are fully reentrant.

@@

One challenge with any DBT system is how it copes with non-reentrant instrumentation. For example, instrumentation that modifies CPU-private data structures cannot be safely executed by a pre-emptive kernel running on an SMP. If the thread being instrumented is interrupted, then the kernel might decide to re-schedule the task to resume its execution on a different CPU. Any modifications to CPU-private state would lead to undefined behavior.

DRK's solution to this problem assumes that all injected instrumentation instructions are potentially non-reentrant, and \emph{delays} interrupts occuring within instrumented code \cite{DRK}. Interrupt delaying allows DRK to maintain the precise interrupt semantics of x86-64: interrupts and exceptions only arrive on logical instruction boundaries in the native code, and kernel interrupt handlers only observe faults or exceptions at native instruction boundaries. This approach introduces subtle complexity when creating instrumentation that injects instructions both before (pre) and after (post) native instructions.

\Cref{fig:interrupt_delay_drk} shows an example of pre- and post-instrumentation around a \texttt{mov} instruction. If an interrupt arrives within the pre-instrumentation, then DRK delays they delivery of the interrupt until the native instruction (\texttt{mov}). After the interrupt is handled and execution attempts to return to the interrupted instructions, DRK will re-instrument the ``tail" of interrupted basic block, starting from the \texttt{mov} instruction. This involves re-creating the pre- and post-instrumentation of the \texttt{mov} in the tail basic block. When executed, this will result in an unbalanced number of \texttt{push} and \texttt{pop} instructions. That is, \texttt{push \%rax} is executed in both the original and tail basic block, but \texttt{pop \%rax} is only executed in the tail basic block. This is a specific example of a more general problem: DRK's approach to interrupt delaying does not play well with idempotent instrumentation.

@@

Btkernel provides no solution to the problem of handling non-reentrant instrumentation, beyond requiring that instrumentation tools disable and re-enable interrupts around non-reentrant instrumentation code \cite{btkernel}. Unfortunately, disabling and re-enabling interrupt is costly.

Granary charts a middle ground between the two aforementioned approaches. Most instrumentation is re-entrant, partially because kernel instrumentation tools can depend on the runtime call stack for scratch storage space. However, interrupt delaying is sometimes useful because it enables non-reentrant instrumentation to avoid to continually disable and re-enable interrupts. For example, Granary's basic block tracing tool writes to a CPU-private buffer containing the sequence of executed basic block program counters. Operating on a CPU-private buffer is more efficient than operating on a global buffer because it avoids cache line contention and locking on an SMP. However, writing to a CPU-private buffer is only safe if the entire write executes on the same CPU. An equivalent tool developed for btkernel---which does not support interrupt delaying---would have to disable and the re-enable interrupts around the write. The is inefficient because it introduces flag stalls into the instruction pipeline \cite{AgnerMicroarchitecture}, and because interrupts arrive relatively infrequently.

@@

Granary's mechanism of looking up branch targets and resolving control flow depends only on CTI policy and target information. This information is either directly embedded in emitted executable code, or in immutable data structures directly referenced by emitted code. This approach ensures that policy information is never lost or corrupted, even in the face of concurrent executions of the same emitted code, arbitrary pre-emption, and arbitrary resumption. Because Granary's translation mechanism is driven by these two inputs, Granary is otherwise stateless\footnote{This is not quite true, as Granary maintains persistent state in the form of the code cache index. However, the code cache index behaves similarly to memoization in a purely functional programming language: it is an optimization, not a necessity.} across context switches.

@@

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
\subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_value\_is\_zero}). When executed, the function call reads its target from an edge data structure (\texttt{jz\_patch\_data}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}
Like btkernel, Granary translates direct control-flow instructions (CTIs) into equivalent jumps into CTI-specific ``edge code" (\Cref{fig:direct_edge_code}). Edge code lives outside the code cache, so as to not pollute the processor's instruction cache \cite{btkernel}.

When control flows through a basic block $B_{pred}$ for the first time, any jumps taken will transfer control to edge code that is specific to the jump taken. Edge code contains only a single indirect \texttt{call} instruction, thus allowing multiple edge codes to occupy a single cache line. The indirect \texttt{call} transfers control to Granary via a function pointer located in an ``edge data structure", which describes how to translate the target basic block, $B_{succ}$. Granary gains control and looks up or translates the target code according instrumentation policy stored in the edge data structure. Then, Granary uses modifies the original CTI in $B_{pred}$ in place so that it transfers control to $B_{succ}$  (\Cref{sec:patching}). Finally, Granary releases control and context switches to $B_{succ}$.

@@

Granary translates indirect CTIs (\texttt{call}, \texttt{jmp}) into their equivalent direct control flow versions, where the direct versions jump to ``in-edge" code. In-edge code loads the indirect jump target (e.g. a specific register or memory location), saves the flags register, and several scratch registers. Finally, the in-edge uses the indirect target as an index into a global, fixed-size hash table of indirect branch targets. The entries of the hash table either point to Granary or to ``out-edge" code (\Cref{fig:ibl}).

@@

Granary's use of out-edge code avoids re-entrancy issues related to indirect control flow that present in DRK and btkernel. Both DRK and btkernel store the target of an indirect CTI in a CPU-private memory location. To complete an indirect control transfer, both systems perform (roughly) the following steps: \begin{enumerate}[itemindent=\parindent]%
\setlength{\itemsep}{2pt}%
\setlength{\parskip}{0pt}%
	\item Save the flags and disable interrupts.
	\item Save registers.
	\item Lookup the next basic block address.
	\item Store that address in a CPU private memory location.
	\item Restore registers.
	\item Restore the flags and interrupt state.
	\item Indirectly jump to the target basic block through the CPU-private location.
\end{enumerate} %
However, the ordering of steps (6) and (7) introduce a reentrancy problem: the access to the CPU-private location is not safe because an interrupt might arrive before the indirect jump in step (7). This unusual design is motivated by the constraint that the entire lookup procedure must save (2) and restore (5) the register state for correctness, but must also save the target pc \emph{somewhere} so that execution can eventually be redirected. Granary's out-edge code avoids this reentrancy issue by having the out edge code itself (in the form of a direct \texttt{jmp}) contain the target pc.

@@

\subsubsection{Ahead-of-time Translation} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

If a basic block ends in either a copy of a native direct \texttt{jmp} or synthesized direct \texttt{jmp}, then Granary will perform ahead-of-time translation and instrumentation of the targeted basic block. The repeated application of translating \texttt{jmp} targets forms traces in Granary's code cache. Granary optimizes traces by directly connecting basic blocks together when there is internal control flow within the blocks of the trace, and by eliding synthesized \texttt{jmp}s between adjacent basic blocks.

Granary's tracing strategy is based on the observation that the most commonly executed path through code follows a straight line. That is, execution is more likely to fall through a conditional CTI than it is to take the conditional CTI. We experimentally verified Granary's tracing strategy by disabling tracing and relying on Granary's performance counters to count how many conditional branches were patched (by their edge code), versus how many synthesized fall-through \texttt{jmp}s were patched.

@@

\subsubsection{Patching and Alignment}\label{sec:patching}
For efficiency, basic blocks in Granary's code cache are directly connected through direct control flow instructions (\texttt{call}, \texttt{jmp}, \texttt{jCC}). That is, if there is a direct branch from one native instruction to another, then there will be a corresponding direct branch between two translated instructions in Granary's code cache.

Granary opportunistically connects two basic blocks together using a direct branch if \begin{inparaenum}[i)]
	\item the branch transfers control to another basic block within the current trace; or
	\item the branch transfers control to an already translated basic block.
\end{inparaenum} However, not all basic blocks can be connected at translation time, therefore Granary employs edge code and hot-code patching (\Cref{fig:direct_edge_code}) to connect basic blocks at runtime.

@@

One approach to change an instruction's alignment is to pad the basic block with additional \texttt{nop} instructions. An alternative when more than a few bytes of padding are needed is to introduce a single control-flow instruction (\texttt{jmp}) that jumps around the padded area. However, adding extra instructions introduces latency into the fetch/decode/execute cycle. Therefore, when possible, Granary uses ignored instruction prefixes to alter an instruction's alignment. For example, Granary aligns patchable conditional jumps using static branch prediction hint prefixes, which are ignored by recent x86-64 processors \cite{AgnerMicroarchitecture}. Finally, most basic blocks are small (i.e. fit in one cache line), therefore, aligning basic blocks on a cache line boundary naturally avoids the potential for a patchable instruction to straddle two cache lines.

@@

\subsubsection{Trace Allocation}\label{sec:trace_alloc}
Arranging for commonly executed sequences of basic blocks to appear adjacent in memory, also called tracing, typically improves the runtime performance of instrumented programs \cite{DynamoRIO}. Typical approaches to tracing require a runtime profiling infrastructure (e.g. recording basic block execution counts). Granary attempts to reap the benefits of tracing without the runtime profiling overhead by opportunistically forming traces using ahead-of-time translation (\Cref{sec:aot}), and by arranging for related basic blocks to be allocated sequentially in memory.

@@

Trace allocation is implemented by recording the memory pool used to allocate a given basic block in that basic block's meta-data. When Granary needs to translate a new basic block, it inspects the predecessor basic block's meta-data and uses the predecessor basic block's memory pool to service allocation requests for the basic block in translation.

@@

\subsubsection{System Calls}

@@

Some instrumentation tools have a limited scope, and therefore do not need to instrument all code. For example, Granary's Guardrail tool is concerned with finding bugs in Linux device drivers \cite{Guardrail}. Guardrail records traces of executed instructions, and analyzes those traces to find memory access and concurrency bugs. Both phases are heavyweight, and so limiting the scope of what is instrumented to what is of interest (module code) results in better performance. Unfortunately, existing systems are unable to change the granularity of instrumentation. Fortunately, Granary allows Guardrail to set the granularity of instrumentation to only instrument a specific module, while leaving the rest of the kernel to execute without overhead.

>>>>

Instrumenting all code (e.g., the whole kernel) imposes unnecessary performance overheads when only a subset of the code is being analyzed (e.g., modules, functions). In practice, we would like to target instrumentation at specific code for efficiency, while also allowing rich and detailed instrumentation of the targeted code. Limiting the scope of what is instrumented is  important because the overhead introduced by using runtime code instrumentation is usually dominated by the instrumentation itself and not the baseline overhead of the runtime system supporting that instrumentation. Also, specializing based on what is instrumented allows instrumentation tools to focus instrumentation on the code of interest. For example, if one wants to instrument RCU read-side critical sections, as is the case with Granary's \toolname{RCUdbg} tool, then instrumenting all code just to filter out and find those critical sections is wasteful.

@@

Another limitation of existing systems is that they are unable to perform efficient runtime code specialization. That is, they are unable to change how code is instrumented based on the context in which that code will run. This is unfortunate because it means that tools wanting to instrument code differently must create their own solution that works around this deficiency. For example, one way that existing tools can specialize code is to generate as many versions of the code as there are potential specializations. When instrumented code is executing, runtime checks continually check and jump to the correct version of the code for the current execution context. However, runtime checks are inefficient, and generating many versions of the code introduces bloat and pollutes the CPU's instruction cache. Granary avoids these issues by supporting context-specific runtime code specialization that generates multiple versions of basic blocks \emph{just-in-time} (JIT), instead ahead-of-time, and without requiring runtime checks.

%Also, the runtime checks needed to multiplex between different versions of instrumented code are orthogonal to the true goals of the instrumentation tool, thus distracting tool-developers from their primary objectives.

@@

One example tool that uses Granary's JIT-based runtime code specialization feature is \toolname{RCUdbg}: Granary's Read-Copy-Update (RCU) debugging tool. \toolname{RCUdbg} catches bugs related to invalid uses of RCU-protected pointers (\Cref{sec:rcudbg}). \toolname{RCUdbg} distinguishes between two kinds of RCU-protected pointers (reader-side and writer-side) and plain old pointers, and three execution contexts (inside of a read-side critical section, inside of a write-side critical section, and in non-RCU-related code). The actions taken when a pointer is de-referenced varies depending on the kind of pointer and the execution context. Code executing in read-side critical sections is specialized to enforce that only plain-old and read-side pointers are accessed, and that all accesses through read-side pointers are memory reads. Write-side critical sections are specialized in a similar way.

@@

Granary translates and instruments kernel binaries one basic block at a time. In Granary, a basic block is a sequence of x86-64 instructions ending in a conditional branch (\texttt{jCC}), \texttt{ret}, or a \texttt{jmp} instruction. Unlike DRK and btkernel, Granary does not terminate basic blocks at \texttt{call} instructions. This enables longer basic blocks and native-like proceccor return address prediction, but makes Granary non-transparent (\Cref{sec:transparency}). Granary takes control of execution starting from ``natural" program entrypoints (e.g. interrupt vectors, system call entrypoints, module initializers). Entrypoint basic blocks  are translated ahead-of-time, and all other basic blocks are translated just-in-time (JIT) as execution ``discovers'' them.

@@

Granary's JIT-based translation approach means that code executing from the code cache may yield control to Granary to request the address of the next basic block to execute. This happens when there is a control flow transfer in a basic block, and the target of that control flow transfer is not already indexed in Granary's code cache. When instrumented code yields to Granary, a ``context switch'' occurs that transfers execution to a CPU-private stack where Granary operates. Granary operates on a CPU-private stack for two reasons. First, if Granary did not switch stacks, then context switches into Granary risk overflowing the runtime call stack. This can happen because kernel stacks are typically small (4K to 8K in size), and because a context switch could happen when the native stack is nearly exhausted. Second, on an SMP, multiple processors can simultaneously execute Granary code (e.g. to translate code that will later execute). In such cases, Granary depends on CPU-private data structures to avoid contention, and the private stack is one such data structure. A context-switch back to the code cache occurs when the next basic block has been found or translated. At this point, instrumented execution continues. Similar to other DBT systems, Granary uses hot code patching (also called code splicing) to reduce the number of context switches. In our experience, kernel code stabilizes very quickly: context switches stop happening after the first few seconds of translation.

@@

Associated with each basic block is meta-data that describes the basic block. Basic block meta-data includes: \begin{itemize}
	\item The length in bytes of the original and translated basic blocks, and the address of the first instruction in the original basic block and the instrumented basic block. This information helps provide a rich debugging interface for instrumentation tool authors. Granary is unique among DBT tools in that the debugging experience is actually pleasant. For example, Granary has a built-in debugging mechanism for stepping back in time through previously executed basic blocks, as well as querying past register state.
	\item Linux kernel-specific exception table meta-data about instructions contained within the current basic block. This kernel-specific meta-data is used by Granary to implement transparent page fault exception handling (\Cref{sec:interrupt_return_address_transparency}).
	\item Ranges of instructions that must be executed atomically. Granary uses this information to implement selective interrupt delaying (\Cref{sec:interrupt_delay}).
	\item The allocator used to allocate the instructions of the basic block. Granary uses this information to implement its trace-allocation optimization (\Cref{sec:trace_alloc}).
	\item The instrumentation policy used to instrument the basic block. Granary uses this information to allow instrumentation tools to define policy-specific interrupt handling routines. For example, \toolname{RCUdbg} implements a read-side policy-specific interrupt handling routine that reports an error if code executing within a read-side critical section is interrupted (\Cref{sec:rcudbg}).
\end{itemize} %
%Granary uses basic block meta-data for trace allocation (\Cref{sec:trace_alloc}), interrupt delaying (\Cref{sec:interrupt_delay}), and policy-specific interrupt handling.

Basic block meta-data can be also be extended by instrumentation tools. This makes developing instrumentation tools that record per-basic block statistics very easy to make. For example, Granary's control-flow graph tool, \toolname{cfg}, extends the basic block meta-data to track per-block and per-edge execution counts (\Cref{sec:cfg}). This data is useful because it provides precise code coverage information (block and branch coverage), as well as useful data for profile-guided optimization. This meta-data is specified by \toolname{cfg} and automatically allocated by Granary.

@@

\subsection{Transparency}\label{sec:transparency}
An instrumentation system is transparent if, given the same inputs, the instrumented and uninstrumented versions of a program behave in the same way \cite{Transparency}. Granary includes configurable levels of transparency to address the trade-off between transparency and overhead. By default, Granary uses a relaxed transparency \cite{btkernel} model for efficiency, flexibility and increased visibility, but if instrumentation for some module requires transparency then it can be enabled at the cost of increased overheads and decreased visibility. By default, Granary exposes the following two artifacts: code cache addresses as function and interrupt return addresses.

@@

\subsubsection{Granary}
For efficiency, most kernel-space instrumentation systems (Granary included) operate on CPU-private state (e.g. Granary's CPU-private stacks). However, accessing CPU-private data without first disabling interrupts results in undefined behavior. Therefore, context-switches into Granary first disable interrupts and then switch execution onto one of Granary's CPU-private stacks. Context switches out of Granary and into native code or the code cache returns execution to the native stack and re-enable interrupts. Execution behavior is preserved because Granary saves the machine state (i.e. registers, flags) in a reentrant way. All other parts of Granary are fully reentrant.

@@

One challenge with any DBT system is how it copes with non-reentrant instrumentation. For example, instrumentation that modifies CPU-private data structures (without disabling interrupts) cannot be safely executed by a pre-emptive kernel running on an SMP. If the thread being instrumented is interrupted, then the kernel might decide to re-schedule the thread to resume its execution on a different CPU. Any modifications to CPU-private state would lead to undefined behavior. An example of this is instrumentation that acquires a CPU-private lock on one CPU, is interrupted, and later releases the wrong lock on the wrong CPU.

DRK's solution to this problem assumes that all injected instrumentation instructions are potentially non-reentrant, and \emph{delays} interrupts occuring within instrumented code \cite{DRK}. Interrupt delaying allows DRK to maintain the precise interrupt semantics of x86-64: interrupts and exceptions only arrive on native instruction boundaries, and kernel interrupt handlers only observe faults or exceptions at native instruction boundaries. This approach introduces subtle complexity when injecting non-idempotent instrumentation before (pre) and after (post) native instructions.

\Cref{fig:interrupt_delay_drk} shows an example of pre- and post-instrumentation around a \texttt{mov} instruction. If an interrupt arrives within the pre-instrumentation, then DRK delays they delivery of the interrupt until the native instruction (\texttt{mov}). After the interrupt is handled and execution attempts to return to the interrupted instructions, DRK will re-instrument the ``tail" of interrupted basic block, starting from the \texttt{mov} instruction. This involves re-creating the pre- and post-instrumentation of the \texttt{mov} in the tail basic block. When executed, this will result in an unbalanced number of \texttt{push} and \texttt{pop} instructions. That is, \texttt{push \%rax} is executed in both the original and tail basic block, but \texttt{pop \%rax} is only executed in the tail basic block. This is a specific example of a more general problem: DRK's approach to interrupt delaying does not play well with non-idempotent instrumentation.

@@

Btkernel provides no solution to the problem of handling non-reentrant instrumentation, beyond requiring that instrumentation tools disable and re-enable interrupts around non-reentrant instrumentation code \cite{btkernel}. The is inefficient because it introduces flag stalls into the instruction pipeline \cite{AgnerMicroarchitecture}, and because interrupts arrive relatively infrequently.

Granary charts a middle ground between the two aforementioned approaches. Most instrumentation is re-entrant, partially because kernel instrumentation tools can depend on the runtime call stack for scratch storage space. However, interrupt delaying is sometimes useful because it enables non-reentrant instrumentation to avoid having to continually disable and re-enable interrupts. For example, Granary's \toolname{Guardrail} tool records traces of executed module instructions into CPU-private buffers \cite{Guardrail}. These traces are concurrently analyzed for bugs by another thread. Operating on a CPU-private buffer is more efficient than operating on a global buffer because it avoids cache line contention and locking on an SMP. However, writing to a CPU-private buffer is only safe if the entire write executes on the same CPU. An equivalent tool developed for btkernel---which does not support interrupt delaying---would have to disable and the re-enable interrupts around the write.

@@

Granary's mechanism of looking up branch targets and resolving control flow depends only on control flow instruction (CTI) policy and target information. This information is either directly embedded in emitted executable code, or in immutable data structures directly referenced by emitted code. This approach ensures that policy information is never lost or corrupted, even in the face of concurrent executions of the same emitted code, arbitrary pre-emption, and arbitrary resumption. Because Granary's translation mechanism is driven by these two inputs, Granary is otherwise stateless\footnote{This is not quite true, as Granary maintains persistent state in the form of the code cache index. However, the code cache index behaves similarly to memoization in a purely functional programming language: it is an optimization, not a necessity.} across context switches.

@@

\begin{figure}[t!]
\subfloat[Native Code]{\usebox\nativejcc}
\hfill
\subfloat[Translated Code]{\usebox\translatedjcc}
\caption{\label{fig:direct_edge_code}Example translation of a conditional control flow instruction (\texttt{jz}) and its edge code. The original \texttt{jz} is translated into a \texttt{jz} that uses a 32-bit relative displacement to transfer control to an edge code function call (\texttt{edge\_value\_is\_zero}). When executed, the function call reads its target from an edge data structure (\texttt{jz\_patch\_data}), which contains all information necessary to resolve the \texttt{jz}'s target and patch the original \texttt{jz}. Patching is performed with a four byte atomic write to the relative displacement operand of the \texttt{jz} instruction. Finally, the return address of the \texttt{call} instruction is altered to transfer control to the newly resolved target.}
\end{figure}
Like btkernel, Granary translates direct CTIs into equivalent jumps into CTI-specific ``edge code" (\Cref{fig:direct_edge_code}). Edge code lives outside the code cache, so as to not pollute the processor's instruction cache \cite{btkernel}.

When control flows through a basic block $B_{pred}$ for the first time, any jumps taken will transfer control to edge code that is specific to the jump taken. Edge code contains only a single indirect \texttt{call} instruction, thus allowing multiple edge codes to occupy a single cache line. The indirect \texttt{call} transfers control to Granary via a function pointer located in an ``edge data structure", which describes how to translate the target basic block, $B_{succ}$. Granary gains control and looks up or translates the target code according instrumentation policy stored in the edge data structure. Then, Granary modifies the original CTI in $B_{pred}$ so that it transfers control to $B_{succ}$  (\Cref{sec:patching}). Finally, Granary releases control and context switches to $B_{succ}$.

@@

Granary translates indirect CTIs (\texttt{call}, \texttt{jmp}) into their equivalent direct control flow versions, where the direct versions branch to ``in-edge" code. In-edge code loads the indirect jump target (e.g. a specific register or memory location), saves the flags register, and several scratch registers. Finally, the in-edge uses the indirect target as an index into a global, fixed-size hash table of indirect branch targets. The entries of the hash table either point to Granary or to ``out-edge" code (\Cref{fig:ibl}).

@@

Granary's use of out-edge code avoids re-entrancy issues related to indirect control flow that are present in DRK and btkernel. Both DRK and btkernel store the target of an indirect CTI in a CPU-private memory location. To complete an indirect control transfer, both systems perform (roughly) the following steps: \begin{enumerate}[itemindent=\parindent]%
\setlength{\itemsep}{2pt}%
\setlength{\parskip}{0pt}%
	\item Save the flags and disable interrupts.
	\item Save registers.
	\item Lookup the next basic block address.
	\item Store that address in a CPU private memory location.
	\item Restore registers.
	\item Restore the flags and interrupt state.
	\item Indirectly jump to the target basic block through the CPU-private location.
\end{enumerate} %
However, the ordering of steps (6) and (7) introduce a reentrancy problem: the access to the CPU-private location is not safe because an interrupt might arrive before the indirect jump in step (7). This unusual design is motivated by the constraint that the entire lookup procedure must save (2) and restore (5) the register state for correctness, but must also save the target pc \emph{somewhere} so that execution can eventually be redirected. Granary's out-edge code avoids this reentrancy issue by having the out edge code itself (in the form of a direct \texttt{jmp}) contain the target pc.

@@

\subsubsection{Ahead-of-time Translation Optimization} \label{sec:aot}
Granary ends basic blocks at conditional branches (\texttt{jCC}), direct and indirect \texttt{jmp}s, and \texttt{ret}s and \texttt{iret}s. Basic blocks ending in conditional branches are extended to contain two trailing CTIs: a copy of the native \texttt{jCC}, and a synthesized direct \texttt{jmp} to the code executed when the condition code is not set (i.e. when the conditional jump is not taken).

If a basic block ends in either a copy of a native direct \texttt{jmp} or synthesized direct \texttt{jmp}, then Granary will perform ahead-of-time translation and instrumentation of the targeted basic block. The repeated application of translating \texttt{jmp} targets forms traces in Granary's code cache. Granary optimizes traces by directly connecting basic blocks together when there is internal control flow within the blocks of the trace, and by eliding \texttt{jmp}s between adjacent basic blocks.

Granary's tracing strategy is based on the observation that the most commonly executed path through code follows a straight line. That is, execution is more likely to fall through a conditional CTI than it is to take the conditional CTI. We experimentally verified Granary's tracing strategy by disabling tracing and relying on Granary's internal performance counters to count how many conditional branches were patched (by their edge code, \texttildelow 25\%), versus how many synthesized fall-through \texttt{jmp}s were patched (\texttildelow 80\%).

@@

\subsubsection{Patching and Alignment  Optimizations}\label{sec:patching}
For efficiency, basic blocks in Granary's code cache are directly connected through direct control flow instructions (\texttt{call}, \texttt{jmp}, \texttt{jCC}). That is, if there is a direct branch from one native instruction to another, then there will be a corresponding direct branch in the translated code.

Granary opportunistically connects two basic blocks together using a direct branch when: \begin{inparaenum}[i)]
	\item the branch transfers control to another basic block within the current trace; or
	\item the branch transfers control to an already translated basic block.
\end{inparaenum} However, not all basic blocks can be connected at translation time, therefore Granary employs edge code and hot-code patching (\Cref{fig:direct_edge_code}) to connect basic blocks at runtime.

@@

One approach to change an instruction's alignment is to pad the basic block with additional \texttt{nop} instructions. An alternative when more than a few bytes of padding are needed is to introduce a single control-flow instruction (\texttt{jmp}) that jumps around the padded area. However, adding extra instructions introduces latency into the processor's fetch/decode/execute cycle. Therefore, when possible, Granary uses ignored instruction prefixes to alter an instruction's alignment. For example, Granary aligns patchable conditional jumps using static branch prediction hint prefixes, which are ignored by recent x86-64 processors \cite{AgnerMicroarchitecture}. Finally, most basic blocks are small (i.e. fit in one cache line), therefore, aligning basic blocks on cache line boundaries naturally avoids the potential for a patchable instruction to straddle two cache lines.

@@

\subsubsection{Trace Allocation Optimization}\label{sec:trace_alloc}
Arranging for commonly executed sequences of basic blocks to appear adjacent in memory, also called tracing, typically improves the runtime performance of instrumented programs \cite{DynamoRIO}. Typical approaches to tracing require a runtime profiling infrastructure \cite{DynamoRIOOptimization}. Granary attempts to reap the benefits of tracing without the runtime profiling overhead by opportunistically forming traces using ahead-of-time translation (\Cref{sec:aot}), and by arranging for related basic blocks to be allocated sequentially in memory.

@@

Trace allocation is implemented by recording the memory pool used to allocate a given basic block in that basic block's meta-data (\Cref{sec:metadata}). When Granary needs to translate a new basic block, it inspects the predecessor basic block's meta-data and uses the predecessor basic block's memory pool to service allocation requests for the basic block in translation.

@@

\subsubsection{System Call Table Optimization}

@@

Some instrumentation tools have a limited scope, and therefore do not need to instrument all code. For example, Granary's \toolname{Guardrail} tool is concerned with finding bugs in Linux device drivers \cite{Guardrail}. \toolname{Guardrail} records traces of executed instructions, and analyzes those traces to find memory access and concurrency bugs. Both phases are heavyweight, and so limiting the scope of what is instrumented to what is of interest (module code) results in better performance. Unfortunately, existing systems are unable to change the granularity of instrumentation. Fortunately, Granary allows \toolname{Guardrail} to set the granularity of instrumentation to only instrument a specific module, while leaving the rest of the kernel to execute without overhead.
